### 2019-11-12 manuscript/manuscript.rmd
revise methods

Taxa sampling plays a critical role in phylogenetic inference, especially for the phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Henjol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ). Importantly, some earlier studies only include less than five ctenophore species. The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position, and it should be noted that Whelan2017 [@whelan2017ctenophore] and Simion2017 [@simion2017large] have much broader taxon sampling than other data matrices: Simion2017 has the most taxon sampling, and Whelan2017 contains greater ctenophore taxon sampling.

@@

### Phylogenetic analyses
#### Full data matrices
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 16 data matrices from recent studies that were constructed from transcriptomic or genomic data (Table). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root of animal (e.g. site heterogeneous CAT model and site homogeneous model), we first investigated four different sets of substitution models in IQtree [@nguyen2014iq], including C10 to C60 profile mixture models as variants of the CAT model in ML framework. Phylogenomic analyses of all the datasets were first conducted under WAG+G, GTR+G, associated best-fit substitution model (C10-C60 model were included for model comparison via -madd option) with ModelFinder [@kalyaanamoorthy2017modelfinder] and Poisson + C60 models using IQtree v [@nguyen2014iq]. Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.

#### Outgroup taxa sampling with C10-C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanozoa only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi. For each Choanozoa data matrice, both C10-C60 models in ML and CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree. Moreover, bayesian inference with the site-heterogeneous CAT-Poisson substitution model was done with PhyloBayes MPI. To minimize computational burden, CAT-GTR models were not in this study.

For results of choanozoa matrices indicated a strong support that sponges are the sister group to the remaining Metazoa using CAT model, bayesian inference with CAT-Poisson model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.  For CAT-poisson analyses on all the data matrices, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. All analyses in PhyloBayes were run for at least one month computational time.

#### phylogenetic signal in conflicted dataset
To investigate the distribution of phylogenetic signal in data matrices, we considered four data matrices from four studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices.We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanozoa and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above. The two different
phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding
tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted by R package.  The numbers of genes and sites supporting each hypothesis were calculated with RAxML output and Perl scripts from Shen et al. For BI analysis, we only considered the Whelan_2017 dataset.

>>>>

Taxa sampling plays a critical role in phylogenetic inference, especially for the phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ). Importantly, some earlier studies only include less than five ctenophore species. The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position, and it should be noted that Whelan2017 [@whelan2017ctenophore] and Simion2017 [@simion2017large] have much broader taxon sampling than other data matrices: Simion2017 has the most taxon sampling, and Whelan2017 contains greater ctenophore taxon sampling.

@@

### Phylogenetic analyses

#### Phylogenetic analyses in IQtree
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root of animal tree of life (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C models in the model testing. For computational efficiency, the GTR with C models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60 and associated best-fit model identified above. Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis. Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C model or CAT site-heterogeneous models. To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models. For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*. 2009 [@hejnol2009assessing], since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept). It should be noted that the main conclusion of Simion et al. 2008 was also based on selection of 25% of genes for their jackknife approach.

#### Outgroup taxa sampling with C10-C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C models IQtree and Poisson-CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree. Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI. To minimize computational burden, CAT-GTR models were not in this study.

For results of choanozoa matrices indicated a strong support that sponges are the sister group to the remaining Metazoa using CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloByes analyses were run for at least one month.

#### phylogenetic signal in conflicted dataset
To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from 3 studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices. We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above. The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted in a customized R script. The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al. 2017 [@shen2017contentious]. For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.
